---
title: "Daily Commute Routines and Adaptability in Sydney (2020–2025)"
author: "Nihira Sharma(530784106), Paakhi Dodwani(SID), Pranav Lokhande(SID)"
date: "2025-07-22"
output:
  html_document:
    toc: true
    toc_depth: 3
    number_sections: true
    code_folding: hide
    theme: united
    highlight: tango
---

```{r}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

```{r}
library(sf)
library(dplyr)
library(ggplot2)
library(ggiraph)
library(rnaturalearth)
library(rnaturalearthdata)
```

# Executive Summary

[TODO]  

- Brief overview of the problem or question
- Why it matters (context, relevance)
- What you aim to discover or demonstrate

# Data Exploration

[TODO]

- Description of datasets used
- Initial exploration and data cleaning steps
- Explanation of analysis methods

# Analysis

[TODO]

## Impact of COVID-19 Lockdowns on Morning Peak Travel (T1 North Shore Line)

[TODO]

- Overview of typical morning peak travel patterns on T1 before 2020
- Description of lockdown timeline and key restrictions in early 2020
- Analysis of ridership decline from January to April 2020 (7–9 AM entries)
- Visualization ideas: line chart of daily entries, spatial map of % drop by station
- Discussion of remote work adoption and mobility restrictions as causes
- Comparison with other lines or broader transport trends (if applicable)
- Implications for future pandemic or emergency transport planning

## Behavioral Shifts During the Delta Wave — Off-Peak Travel on T3 Bankstown

[TODO]

- Context on Delta wave timing and public health advice in 2021
- Explanation of peak vs off-peak travel definitions and typical patterns pre-Delta
- Analysis of hourly entry data showing flattening of morning peaks and off-peak increases
- Visualization ideas: heatmap or stacked bar chart for hourly usage; clustered map of stations with strongest shifts
- Possible reasons: crowd avoidance, flexible work hours, staggered shifts
- Comparison to T1 and other lines to highlight unique or common behaviors
- Consideration of whether shifts persisted beyond Delta wave period

## Weather Impact — Rainy Season Effects on T2 Inner West Line

[TODO]

- Description of the 2022 prolonged rainy season and key heavy rainfall events
- Data sources: rainfall measurements, daily ridership numbers
- Correlation analysis between rainfall intensity and ridership drops on T2 line
- Visualization ideas: dual-axis line chart overlaying rainfall and train entries
- Discussion of weather-related disruptions on commuter reliability and modal shifts
- Implications for infrastructure resilience and transport scheduling under adverse weather

## Return to Office — Commute Frequency on T4 Eastern Suburbs Line in 2023

[TODO]

- Overview of easing restrictions and return-to-office policies starting in 2023
- Analysis of morning peak ridership recovery trends on T4 line compared to pre-pandemic levels
- Visualization ideas: smooth area chart or boxplot showing multi-year commute frequency changes
- Discussion of hybrid work impacts, partial recovery, and remaining gaps in usage
- Comparison to other lines or modes (if data available)
- Potential implications for service planning and future demand forecasting

# Conclusion

 [TODO]
 
 - Summarize main insights
 - Discuss implications or potential next steps
 - Reflect on limitations and what could be done further